---
title: 'Reproducible Research: Course Project 2'
author: "Martijn van Gorp"
date: "13 April 2016"
output: html_document
---
***
## Basic Analysis on the Impact of Severe Weather Events

#### Synopsis

#### Data processing

##### Download file
```{r}
dataurl = 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
datafile = 'stormdata.csv.bz2'

# Check if already present
if(!file.exists(datafile)) { download.file(dataurl,datafile) }
```

##### Read the data
The function read.csv can read bz2 files, so no need to unzip
```{r cache=TRUE}
data = read.csv(datafile, header = TRUE, sep=",")

# Show first few rows to get an idea of the raw dataset
head(data)
```

##### Load libraries
We use the following libraries for this analysis:
```{r}
suppressMessages(library(dplyr))
suppressMessages(library(reshape2))
suppressMessages(library(ggplot2))
```

#### Results
Both research questions are related to the statistics per weather event, 
so we can group the data by weather event type. The event type can be found in de columns EVTYPE.
```{r}
# Group by event
subdata_by_event = data %>% group_by(EVTYPE)
```

We are now prepared to answer the two research questions.

#####Question 1. Across the United States, which types of events are most harmful with respect to population health?
First we have to ask ourselves how we can measure population health. 
If we look at the data set, there are two columns that give the fatalities and injuries for each measured weather event.
So if both of these have a high number, we can say that an event has a harmfull effect on population health.
To give an answer to which events are most harmful, we work with a Top 10.

With the code below, we to calculate the SUM of the columns FATALITIES and INJURIES. 
Next to this we calculate the total of both of these so we can sort the result by this total to get a Top 10.
```{r}
health = subdata_by_event %>% 
    summarise_each(funs(sum), FATALITIES, INJURIES) %>% 
    mutate(TOTAL = (FATALITIES + INJURIES)) %>%
    arrange(desc(TOTAL)) %>%
    top_n(10)

# Show data
health
```

```{r fig.width=10, fig.height=4}
# Sort data by TOTAL (needed for graph)
health$EVTYPE <- factor(health$EVTYPE, levels = health$EVTYPE[order(health$TOTAL)])

# Make wide tabel into long (remove TOTAL column)
health_long = melt(select(health,-TOTAL), id = c("EVTYPE"))

# Barplot
ggplot(health_long, aes(x = EVTYPE, y = value, fill = variable)) + 
        geom_bar(stat='identity') + 
        xlab('Weather Event') + 
        ylab('Number of faculties and injuries') +
        ggtitle('Top 10 Weather Events most harmful to population health') +
        scale_fill_discrete(name = 'Type') +
        coord_flip()
```

#####Question 2. Across the United States, which types of events have the greatest economic consequences?
```{r}
econom = subdata_by_event %>% 
    summarise_each(funs(sum), PROPDMG, CROPDMG) %>% 
    mutate(TOTAL = (PROPDMG + CROPDMG)) %>%
    arrange(desc(TOTAL)) %>%
    top_n(10)

# Show data
econom
```

```{r fig.width=10, fig.height=4}
# Sort data by TOTAL (needed for graph)
econom$EVTYPE <- factor(econom$EVTYPE, levels = econom$EVTYPE[order(econom$TOTAL)])

# Make wide tabel into long (remove TOTAL column)
econom_long = melt(select(econom,-TOTAL), id = c("EVTYPE"))

# Barplot
ggplot(econom_long, aes(x = EVTYPE, y = value, fill = variable)) + 
        geom_bar(stat='identity') + 
        xlab('Weather Event') + 
        ylab('Damage in Dollars') +
        ggtitle('Top 10 Weather Events with greatest economic consequences') +
        scale_fill_discrete(name = 'Type') +
        coord_flip()
```